---
title: "Exploratory_Data_Analysis_2"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "T:/MUGISHA TIMOTHY NAABAASA/YEAR 2 SEMESTER 1/Big Data Analytics and Technologies/Assignments/Main Practical Folder/Notebooks")
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
file.exists("DiamondPricesData.xlsx")
```
#Install and load necessary libraries

```{r install missing necessary libraries}
required_packages <- c("readxl", "dplyr", "tidyr","ggplot2", "corrplot", "gridExtra")

missing_packages <- setdiff(required_packages,rownames(installed.packages()))

if (length(missing_packages) > 0) {
  install.packages(missing_packages)
}
```

```{r global installation of tidyr package}
install.packages(c("purrr", "tidyr"), lib = "C:/Program Files/R/R-4.6/library")

```
```{r}
remove.packages("tibble")
```


```{r}
install.packages("tibble")
```


```{r}
library(tibble)
```


```{r load libraries}
# Read Excel files
library(readxl)

# Manipulate and transform data frames
library(dplyr)

# Reshape data for faceted graphs
library(tidyr)

# Create graphs
library(ggplot2)

# Draw a correlation heatmap
library(corrplot)

# Arrange several ggplot graphs in one figure
library(gridExtra)

```

#Import the dataset

```{r read excel workbook}
data <- read_excel("DiamondPricesData.xlsx")

```

#Inspection of the Data

```{r}
data
```

```{r}
dim(data)
```

```{r}
str(data)
```

```{r}
sapply(data, class)
#sapply() applies a function repeatedly. Here the dataset is data and the function is class, so the command asks R to report the class of each column.
```

#Change categorical variables to factors

```{r}
data <- data %>%mutate(across(c(cut, color, clarity),as.factor))

#%>% is the pipe operator and can be read as “then”. mutate() creates or modifies columns. across() applies the same operation to several selected columns. as.factor() converts a variable to an R factor so R treats it as categorical.
```

```{r confirm conversion}
sapply(data, class)

```

#Summarize the dataset and category levels

```{r}
summary(data)

#summary() reports descriptive statistics for numeric columns and category counts for factor columns.
```

```{r}
lapply(data[c("cut", "color", "clarity")],levels)

#lapply(..., levels) displays the category levels of each selected factor.
```

#Identify and remove duplicate values

```{r}
sum(duplicated(data))
```

```{r}
data_new <- distinct(data)

#distinct() comes from dplyr and keeps unique rows.
```

```{r}
cat("Shape after deleting duplicate values:", nrow(data_new), "rows,", ncol(data_new), "columns\n")

#cat() only prints a readable message; it does not perform the duplicate removal. nrow() counts rows and ncol() counts columns.
```

```{r}
# Shape before removing duplicates
dim(data)

# Shape after removing duplicates
dim(data_new)
```

#Separate categorical and continuous variables

```{r}
cat <- data_new %>% select(where(is.factor))

head(cat)
```

```{r}
cont <- data_new %>% select(where(is.numeric))

head(cont)
```

**Analytical Problem**: We want to predict price using the other variable predictors (independent variables). Because the dependent variable, price, is a continuous variable, we shall use supervised regression.


#Distribution of the target variable

```{r}
ggplot(data_new, aes(x = price)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +    # Divide the observed price range into 30 bins
  labs(title = "Distribution of diamond price", x = "Price", y = "Count") +     # Add a title and axis labels
  theme_minimal()

```
**Interpretation**: This graph describes price as heavily right-skewed: many diamonds occur at lower prices, with a long tail extending to much more expensive diamonds. Thus the data is not normally distributed.

```{r}
summary(data_new)
```
**Interpretation**: This summary alerts us that there's more work to be done, especially with the dealing of outliers and missing values. These must be taken care of before we visualize for insights.



#Compare distributions of categorical and numerical variables

```{r categorical}

plot_bar_charts <- function(data, cols) {
  plots <- lapply(cols, function(col) {
  ggplot(data, aes(x = .data[[col]])) +

  geom_bar(fill = "forestgreen") +

  labs(title = col) +
  theme_minimal()
  })
  
  grid.arrange(grobs = plots, nrow = 1)
  }
  
  plot_bar_charts(
  data_new,
  c("cut", "color", "clarity")
  )
```
**Interpretation**:


```{r}

data_new %>% select(carat, depth, table, x, y, z) %>% tidyr::pivot_longer(everything(), names_to = "variable",
  values_to = "value") %>%

  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "violet", color = "white") +
  facet_wrap(~variable, scales = "free") +
  theme_minimal()
```
**Interpretation**:


#Investigate and Handle Outliers
    In this activity, we will use the low-end and high-end replacement technique. Low-end and high-end replacement is an outlier-handling technique where, instead of deleting extreme observations, you replace values that are too low or too high with reasonable boundary values.
    It is closely related to winsorization, but the terminology can vary depending on the course or implementation:
    
Low-end replacement → extreme low values are replaced.
High-end replacement → extreme high values are replaced.

```{r}
# Function for low- and high-end replacement
replace_outliers <- function(x) {
  
  # Calculate Q1 and Q3
  Q1 <- quantile(x, 0.25, na.rm = TRUE)
  Q3 <- quantile(x, 0.75, na.rm = TRUE)
  
  # Calculate IQR
  IQR_value <- Q3 - Q1
  
  # Calculate boundaries
  lower <- Q1 - 1.5 * IQR_value
  upper <- Q3 + 1.5 * IQR_value
  
  # Low-end replacement
  x[x < lower] <- lower
  
  # High-end replacement
  x[x > upper] <- upper
  
  return(x)
}


# Replace the values while keeping the original variable names
data_new$x <- replace_outliers(data_new$x)
data_new$y <- replace_outliers(data_new$y)
data_new$z <- replace_outliers(data_new$z)

```

```{r Visualize results}
data_new %>% select(x, y, z) %>% tidyr::pivot_longer(everything(), names_to = "variable",
  values_to = "value") %>%

  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "tan", color = "white") +
  facet_wrap(~variable, scales = "free") +
  theme_minimal()
```

**Observation**:


#Identify and Handle missing data
```{r}
colSums(is.na(data_new))
```


```{r}
data_new$depth[is.na(data_new$depth)] <- median(data_new$depth,na.rm = TRUE)
```


```{r}
colSums(is.na(data_new))
```

##EXPLORATION OF RELATIONSHIPS WITH VISUALIZATIONS AND TESTS

#1. Continuous Predictors against Price
```{r}
# Create a vector containing the continuous predictors

continuous_cols <- c("carat", "depth", "table", "x", "y", "z")
```


```{r}
#Generate one scatter plot per predictor

for (predictor in continuous_cols) {
  p <- ggplot(data_new,aes(x = .data[[predictor]],y = price)) +

  geom_point(alpha = 0.15, color = "steelblue") +        # alpha makes dense point clouds more transparent

  labs(title = paste(predictor, "vs price")) +
  theme_minimal()

  print(p)
}
```

**Interpretation**: An increasing pattern means there is a positive association, a negative pattern shows negative association, and no obvious pattern indicates weak/no visible linear association. The drawn graphs report strong increasing trends for carat, x, y, and z, while depth and table show little visible relationship with price.


#Remove extreme dimension rows
```{r}
data_filter <-data_new$z > 2.06 & data_new$z < 6.5
```


```{r}
# Keep rows for which data_filter is TRUE
data_new <- data_new[data_filter, ]

# Count the remaining observations
nrow(data_new)
```


```{r replot}
for (predictor in continuous_cols) {
  p <- ggplot(data_new,aes(x = .data[[predictor]],y = price)) +
  
  geom_point(alpha = 0.15,color = "tomato") +
  
  labs(title = paste(predictor,"vs price (after outlier removal)")) +
  
  theme_minimal()
  print(p)
  }
```

#Measure the strength and direction of the predictors with price

```{r Select Numerical Predictors, plus target}
correlation_cols <- c("price", "carat", "depth","table", "x", "y", "z")
```


```{r Calculate correlation Matrix}
correlation_data <- cor(data_new[, correlation_cols])
```
*Note*: cor() calculates Pearson correlation by default. The result is a square matrix in which each cell contains the correlation coefficient for a pair of numeric variables.


```{r Draw correlation heatmap(upper triangle)}
corrplot(correlation_data,method = "color",type = "upper", addCoef.col = "black",number.cex = 0.8,
  title = "Correlation matrix of diamond variables",mar = c(0, 0, 2, 0))
```
*Note*: "method = "color"" represents correlations using a color scale; type = "upper" displays only the upper triangular half; "addCoef.col" adds the numeric coefficients.

```{r Extract correlations with price}
# Take the price column from the correlation matrix
price_corr <- correlation_data[, "price"]

sort(price_corr[abs(price_corr) > 0.5],decreasing = TRUE)   
# Keep only variables whose absolute correlation with price exceeds 0.5
# Sort the retained correlations from largest to smallest
```

#Explore categorical predictors
```{r}
categorical_cols <- c("cut", "color", "clarity")
```


```{r}
box_plots <- lapply(categorical_cols,function(col) {
  ggplot(data_new,aes(x = .data[[col]],y = price)) +
  
  geom_boxplot(fill = "yellow") + labs(title = paste("price by", col)) +
  
  theme_minimal()
  }
  )

# Place the three boxplots in a single row
grid.arrange(
grobs = box_plots,
nrow = 1
)
```
*Note*: A firmal significance test is still required to deduce whoch category associates with the target varibale

*Hypotheses*
- H0 (null hypothesis): the mean price is the same across all categories of the predictor. 

- HA (alternative hypothesis): at least one category has a different mean price.

- Decision rule: If p-value < 0.05, reject H0 and treat the predictor as statistically associated with price for this practical. If p-value ≥ 0.05, fail to reject H0.
```{r Test using ANOVA}
function_anova <- function(data,target_variable,categorical_predictors) {

  selected_predictors <- c()                    # Start with an empty vector for variables that pass the test
  cat("##### ANOVA Results #####\n\n")
 
  for (predictor in categorical_predictors) {
    formula <- as.formula(paste(target_variable, "~", predictor))       # Build a formula such as: price ~ cut
    anova_result <- summary(aov(formula, data = data))                  # Fit the ANOVA model and obtain its summary table

    p_value <-
    anova_result[[1]][["Pr(>F)"]][1]                                    # Extract the first p-value from the ANOVA table
    

    if (p_value < 0.05) {                                               # Apply the 0.05 decision rule
      cat(predictor,"is correlated with",target_variable,"| P-Value:",p_value,"\n")
      
      selected_predictors <- c(selected_predictors,predictor)           # Add the significant predictor to the selection vector
      } 
    else {
      cat(predictor,"is NOT correlated with",target_variable,"| P-Value:",p_value,"\n")
      }
      }
  
  
  return(selected_predictors)                                         # Return the predictors that passed the significance rule
  }

```


```{r Run the function}
categorical_predictor_list <- c("cut", "color", "clarity")

# Test each categorical predictor against price
selected_categorical <- function_anova(data_new,"price",categorical_predictor_list)

selected_categorical

```

# Combine the continuous variables retained by correlation screening with the categorical variables retained by ANOVA.
```{r Define selected predictor names}
selected_columns <- c("carat", "x", "y", "z","cut", "color", "clarity")
```


```{r Create a predictor-only dataset}
data_for_ml <- data_new[, selected_columns]

head(data_for_ml)
```

**Conclusion**: This practical excludes depth and table because they fall below the chosen correlation threshold with price. The seven retained columns are presented as the candidate predictor set for later regression modelling.


#Summary
```{r}
# Confirm the final number of observations and variables
dim(data_new)

# Confirm that no missing values remain
colSums(is.na(data_new))

# Inspect the selected predictor dataset
str(data_for_ml)
head(data_for_ml)
```

